Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula

### 2024-02-05 output/paper/manuscript.qmd
Started revision manuscript from MV. New csl bib.

bibliography: bibliography.bib
csl: ama11.csl

@@

@@

Results of epidemiological studies seem to indicate that exposure to certain environmental chemicals, including endocrine disruptors (EDCs), is linked to perturbation of the hypothalamic-pituitary-adrenocortical (HPA) axis, which has a major role in brain development. We evaluated the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and the Attention Network Test (ANT) in childhood, using data from the Human Early Life Exposome (HELIX) Study. We used the parametric g-formula and marginal contrasts to estimate effects between EDCs, glucocorticosteroids, and a measure of efficiency from the ANT, the hit reaction time standard error (HRT-SE), in roughly 1,000 children aged 6-11. We further tested for possible effect modification by sex. We found positive marginal contrasts for exposure to the phenol MEPA and the phthalate metabolites oxo-MiNP, oh-MiNP, and MEHP, on HRT-SE. Multiple EDCs were also associated with positive marginal contrasts of cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on HRT-SE, although we found possible effect modification by sex for cortisone and corticosterone production. Our results suggest that multiple EDCs might interfere with the homeostasis of the HPA axis. Further research is necessary to determine whether this can have a clinically significant impact on childhood neurodevelopment.

@@

- Authorship must follow the ICMJE's criteria. Author names should be listed in ScholarOne and author contributions should be detailed in the cover letter ("Author A..."). Neither names nor contributions should appear in the blinded manuscript. Do not include conflicts of interest.
- The length of the abstract must be less or equal than 200 words, and should be unstructured, stating the research questions, the methods used, and the results and conclusions of the research.
- The length must be <4,000 words, excluding abstract, references, tables, figure captions, acknowledgments, and Supplementary Material.
- In-text citations: full-sized Arabic numerals in parentheses within the sentence.
- Submission: SER members receive a 10% discount on the fees per page. This should be requested in the cover letter. Also 20% discount for Open Access charges. If uploading a single file (document, tables, figures, SM), designate the file as "Main document - anonymous".

The prevalence of several neurodevelopmental disorders has increased in the pediatric population [@GrandjeanLandrigan:2014], and multiple environmental pollutants play a role in the increased rates of these disorders [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. \acr{edcs}, chemicals that are capable of interfering with the endocrine system, have been shown to have a role in childhood neurodevelopment and behavior [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @TewarAuingerBraun:2016; @YuDuChiou:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @FurlongHerringBuckley:2017; @Braun:2017; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019; @ShoaffCoullWeuve:2020; @OhKimKannan:2023; @VilmandBeckBilenberg:2023]. Although both pregnancy and early childhood are crucial stages of (neuro)development, most of the available literature is focused on the effects of prenatal exposure to \acr{edcs} on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Moreover, little is known about the biological mechanisms of action [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

\acr{opps} might have a deleterious effect on neurodevelopment, although the few studies conducted on children and using a biological matrix for exposure assessment suffered from a series of limitations, including a small sample size and few \acr{opp} metabolites measured [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Phthalates and their metabolites have been associated to several adverse effects on neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Similarly, these studies focused on few phthalates metabolites and were conducted in small populations. On the other hand, the effects of exposure to \acr{bpa} on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

There is also evidence that exposure to certain \acr{edcs}, specifically phthalates, might interfere with the \acr{hpa} axis and might interact with the glucocorticoid receptor [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023].

@@

targets::tar_load(
  dplyr::matches("marginal_comparisons_(\\d)$"),
  store = store
)
meffects1 <- tidy_res_meffects(
  df = marginal_comparisons_1,
  sa_var = NULL,
  outcome = c("hitrtse"),
  which_res = "comparisons",
  num_digits_est = num_digits_est,
  num_digits_sig = num_digits_sig
)$numerical_results
meffects2 <- tidy_res_meffects(
  df = marginal_comparisons_2,
  sa_var = NULL,
  outcome = c(
    "cortisol production",
    "cortisone production",
    "corticosterone production"
  ),
  which_res = "comparisons",
  num_digits_est = num_digits_est,
  num_digits_sig = num_digits_sig
)$numerical_results
meffects3 <- tidy_res_meffects(
  df = marginal_comparisons_3,
  sa_var = NULL,
  outcome = c("hitrtse"),
  which_res = "comparisons",
  num_digits_est = num_digits_est,
  num_digits_sig = num_digits_sig
)$numerical_results
```

>>>>

bibliography: bibliography.bib
csl: american-journal-of-epidemiology.csl

@@

```{r}
targets::tar_load(
  dplyr::matches("marginal_comparisons_(\\d)$"),
  store = store
)
meffects1 <- tidy_res_meffects(
  df = marginal_comparisons_1,
  sa_var = NULL,
  outcome = c("hitrtse"),
  which_res = "comparisons",
  num_digits_est = num_digits_est,
  num_digits_sig = num_digits_sig
)$numerical_results
meffects2 <- tidy_res_meffects(
  df = marginal_comparisons_2,
  sa_var = NULL,
  outcome = c(
    "cortisol production",
    "cortisone production",
    "corticosterone production"
  ),
  which_res = "comparisons",
  num_digits_est = num_digits_est,
  num_digits_sig = num_digits_sig
)$numerical_results
meffects3 <- tidy_res_meffects(
  df = marginal_comparisons_3,
  sa_var = NULL,
  outcome = c("hitrtse"),
  which_res = "comparisons",
  num_digits_est = num_digits_est,
  num_digits_sig = num_digits_sig
)$numerical_results
```

@@

Epidemiological evidence suggests that exposure to \acr{edcs} may be linked to perturbation of the \acr{hpa} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention measures in children. We used data from the \acr{helix} cohort including 1,297 children aged 6-11. We used the parametric g-formula and \acr{mcs} to estimate effects between \acr{edcs}, glucocorticosteroids, and \acr{hitrtse}, a measure of efficiency from the \acr{ant}, and tested for possible effect modification by sex. We observed positive \acr{mcs} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "hitrtse")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "hitrtse")`), oh-MiNP (`r format_res(meffects1, "ohminp", "hitrtse")`), and MEHP (`r format_res(meffects1, "mehp", "hitrtse")`), on \acr{hitrtse}. Several \acr{edcs} were also associated with positive \acr{mcs} of cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on \acr{hitrtse}, although we found possible effect modification by sex for cortisone and corticosterone production. Our results suggest that multiple \acr{edcs} might interfere with inattentiveness outcomes in children and with the homeostasis of the \acr{hpa} axis. Further research is necessary to determine whether this can have a clinically significant impact on childhood neurodevelopment.

@@

- Authorship must follow the ICMJE's criteria. Author names should be listed in ScholarOne and author contributions should be detailed in the cover letter ("Author A..."). Neither names nor contributions should appear in the blinded manuscript. Do not include conflicts of interest.
- The length of the abstract must be less or equal than 200 words, and should be unstructured, stating the research questions, the methods used, and the results and conclusions of the research.
- The length must be <4,000 words, excluding abstract, references, tables, figure captions, acknowledgments, and Supplementary Material.
- Submission: SER members receive a 10% discount on the fees per page. This should be requested in the cover letter. Also 20% discount for Open Access charges. If uploading a single file (document, tables, figures, SM), designate the file as "Main document - anonymous".

The prevalence of several neurodevelopmental disorders has increased in the pediatric population [@GrandjeanLandrigan:2014], and multiple environmental pollutants may play a role in the increased rates of these disorders [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. \acr{edcs}, ubiquitous chemicals present in many every-day products and diet, are capable of interfering with the endocrine system, and have shown associations with childhood neurodevelopment and behavior [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @TewarAuingerBraun:2016; @YuDuChiou:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @FurlongHerringBuckley:2017; @Braun:2017; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019; @ShoaffCoullWeuve:2020; @OhKimKannan:2023; @VilmandBeckBilenberg:2023]. Although both pregnancy and early childhood are crucial stages of (neuro)development, most of the available literature is focused on the effects of prenatal exposure to \acr{edcs} on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

\acr{opps} are one group of \acr{edcs} that may have a deleterious effect on neurodevelopment, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size and few \acr{opp} metabolites measured [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites has been associated to several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to \acr{bpa} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

Moreover, little is known about the biological mechanisms of action [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].
There is also evidence that exposure to certain \acr{edcs}, specifically phthalates, might interfere with the \acr{hpa} axis and might interact with the glucocorticoid receptor [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023].

@@
